Memes: an R interface to the MEME Suite

### 2020-10-30 manuscript.Rmd
Finish second draft edits.

Biopolymers, such as DNA and protein, perform varying functions based on their
primary sequence. Short, repeated sequences, or "motifs" represent functional
units within biopolymers which can act as interaction surfaces, create
structure, or have enzymatic activity. Identification of shared motifs across
multiple sequences can provide evidence for shared function, for example,
identification of kinase substrates based on similarities in phosphorylation
site sequence, or characterizing DNA elements based on shared transcription
factor binding sequences (TODO: ref kinase motif:
https://doi.org/10.1016/0968-0004(90)90073-K). Thus the ability to
identify, classify, and compare motifs represents a key step in the analysis of
biological sequences. **TODO: R/Bioconductor transition sentence**. Although
tools that implement these features have historically existed in the
Bioconductor landscape, they often rely on proprietary data types with poor
interoperability with other tools, creating barriers to comprehensive motif
analysis within R. 
**TODO: fix this section, how to transition?**
Furthermore, popular motif analysis packages, MotIV, and motifRG 
will be deprecated in the next Bioconductor release,
leaving a gap in functionality with no clear alternatives.

@@

In addition to wrapping MEME Suite utilities, `memes` extends the functionality
of `universalmotif` to provide base R and `tidyverse` compatible syntax for
manipulating and analyzing motif metadata. These features allow users to
integrate external data with motif databases to aid analysis, such as filtering
a set of transcription factor motifs based on gene expression in the tissue of
interest, then using that filtered set to search for motifs associated with
expressed factors.

@@

 - as mentioned previously, the multidimensional naature of meme results is difficult to represent
 - Output data usually contain motifs & descriptive statistics about those motifs
 - motif info can be stored in universalmotif format, however, not enough space for data
 - data.frames do good job of storing data (incl nested data), but cannot store motifs
 - we solve this problem by storing the unviversalmotif inside the data.frame (which is valid R)
 - we add nested list columns as needed to link multidimensional data
 - advantage of this approach over an S4 is all normal tools work on it (tidyverse), joins, etc.

Many of the MEME Suite tools return files contianing descriptive
properties of motifs. For example, MEME and DREME return de-novo discovered
motifs from query sequences along with statistical information about their
enrichment. The `universalmotif` data structure is a powerful R/Bioconductor
representation for motif matrices and their associated metadata
\cite{universalmotif}.  However, `universalmotif` objects are not ideal for
performing bulk manipulation or queries of metadata, nor do they support complex
data structures such as nested elements. Many MEME Suite analysis pipelines
involve identifying motifs which are further analyzed by multiple tools,
creating additional metadata which cannot be easily tracked in bulk using
`universalmotif` format. Thus an ideal representation for these data is one
which can hold an unlimited amount of metadata, contain complex hierarchies, and
be easily manipulated using standard analysis tools. `memes` therefore extends
the `universalmotif` format by providing data.frames holding motif metadata
which retain links to their `universalmotif` representation. These
universalmotif data.frames combine the advantages of the `universalmotif`
framework for motif manipulation with the extensibility and flexibility of R
data.frames. These `universalmotif` data.frames form the basis for a majority of
`memes` data outputs. These structures are also valid input types to `memes`
functions, and when used as such, output data are appended as new columns to the
input data, ensuring data provenance.

@@

`universalmotif` data.frames are not limited to storing motifs from MEME Suite
data, and can additionally be used to examine or manipulate the properties of
any motif list or database. Using `universalmotif` functions to import motif data,
the `memes` function `as_universalmotif_dataframe()` converts any `universalmotif` object
into a `universalmotif` data.frame which can be manipulated using base R or
`tidyverse` functions, just like regular R data.frames. Manipulation of linked
metadata columns will result in updating the linked `universalmotif` object when
calling `update_motifs()` to retain the data.frame structure, or when calling
`as_universalmotif()` to convert the data.frame back into `universalmotif` format. In
this way, `memes` enables large-scale motif database manipulation & analysis
within the preexisting R framework.

@@

Motif analysis is often employed in ChIP-seq analysis, in which data is stored
as genomic coordinates rather than sequence. However, MEME Suite tools are
designed to work with sequences. While existing tools such as bedtools can
extract DNA sequence from gneomic coordinates, some MEME tools require fasta
headers to be specifically formatted. As a result, users must write custom code
to extracd DNA sequence for their genomic ranges of interest.

@@

During analysis of multiomic data, users can identify different functional
categories of genomic regions through integration with orthogonal datasets. **TODO
For example, .**

In these instances, motif analysis acts as a powerful tool to
examine whether differences in motif content is associated with membership of
distinct categories. In some cases, performing differential motif analysis
(using one category as the "test" set and another as the "control") can uncover
biologically relevant motifs which are enriched in one category relative to
another. Although the MEME Suite allows differential enrichment testing, it does
not inherently provide a mechanism for analyzing groups of sequences in
parallel, or performing motif analysis with any understanding of data
categories. The `memes` framework enables "data-aware" motif analysis workflows
by allowing named lists of `Biostrings` objects as input to each function. If
using `GenomicRanges`, users can split peak data on a metadata column using the
base R `split` function, and then use this result as input to `get_sequence`, which
will produce a list of `BioStrings` objects where each entry is named after the
data categories from the split column. When a list is used as input to a `memes`
function, it runs the corresponding MEME Suite tool for each object in the list.
Users can also pass the name(s) of a category to the `control` argument to
enable differential analysis of the remaining list members against the
control category sequences. Therefore, `memes` enables complex data-aware
differential motif analysis workflows using simple syntax to extend the
capabilities of the MEME Suite.

@@

The TomTom tool allows users to compare unknown motifs to a set of known motifs
to identify the best match. Visual inspection of motif comparison data is a key
step in assessing the quality and accuracy of a match. The `view_tomtom_hits`
function allows users to compare query motifs with the list of potential matches
as assigned by TomTom. The `force_best_match` function allows users to
reassign the TomTom best match to a lower-ranked (but still statistically
significant) match in order to highlight motifs with greater biological
relevance (e.g. to skip over a transcription factor that is not expressed in the
experimental sample, or when two motifs are matched equally well).

>>>>

Biopolymers, such as DNA and protein, perform varying functions based on their
primary sequence. Short, repeated sequences, or "motifs" represent functional
units within biopolymers which can act as interaction surfaces, create
structure, or have enzymatic activity. Identification of shared motifs across
multiple sequences can provide evidence for shared function, for example,
identification of kinase substrates based on similarities in phosphorylation
site sequence, or characterizing DNA elements based on shared transcription
factor binding sequences (TODO: ref kinase motif:
https://doi.org/10.1016/0968-0004(90)90073-K). Thus the ability to identify,
classify, and compare motifs represents a key step in the analysis of biological
sequences. **TODO: R/Bioconductor transition sentence**. Although tools that
implement these features exist in Bioconductor, they often rely on
proprietary data types or have poor interoperability with other tools, creating
barriers to comprehensive motif analysis within R. The recent addition of
`universalmotif` to the Bioconductor landscape partially addresss these issues
by standardizing motif formats and providing a powerful set of tools for
examining the properties of motifs, however the Bioconductor motif analysis
landscape remains incomplete \cite{universalmotif}.

@@

In addition to wrapping MEME Suite utilities, `memes` extends the functionality
of `universalmotif` to provide base R and `tidyverse` compatible syntax for
manipulating and analyzing motif metadata. These features allow users to
integrate external data with motif databases to aid analysis, such as filtering
a set of transcription factor motifs based on gene expression in the tissue of
interest, then using that filtered set to search for motifs associated with
expressed factors. By maintaining compatability with `universalmotif`, `memes`
outputs can be readily used in a diverse set of tools beyond the `memes`
framework.

@@

Data returned from MEME Suite tools are complex and multidimensional, making
them difficult to represent in a simple data structure. For example, MEME and
DREME return de-novo discovered motifs from query sequences along with
statistical information about their enrichment. In this instance, a position
frequency matrix (PFM) of the discovered motif can be represented as a matrix,
while the properties of that matrix (e.g. the name of the motif, the E-value
from the enrichment test, background nucleotide frequency, etc.) must be encoded
outside of the matrix, while maintaining their relationship to the
corresponding PFM. However, other MEME Suite tools, such as TOMTOM, which
compares one motif to a series of several motifs to identify possible matches,
produce an additional layer of complexity such that input matrices (which
contain metadata as previously described) can have a one-to-many relationship
with other motif matrices (again with their own metadata). Thus an ideal
representation for these data is one which can hold an unlimited number of motif
matrices and their metadata, contain complex hierarchies, and be easily
manipulated using standard analysis tools.

The `universalmotif` data structure is a powerful R/Bioconductor representation
for motif matrices and their associated metadata \cite{universalmotif}. However,
`universalmotif` objects are not ideal for performing bulk manipulation or
queries of metadata, nor do they support complex data structures such as nested
elements. R data.frames support nesting arbitrary data types within columns, but
are not sufficient to properly store motif matrices. In order to create an
extensible, nested data structure for tracking MEME Suite data, `memes`
leverages the power of R data.frames and the `universalmotif` format by nesting
`universalmotif` objects within data.frames. These universalmotif data.frames
store information about motifs along rows, where each column stores metadata
associated with a motif, including a `motif` column holding a `universalmotif`
representation of that motif, which can be manipulated by altering columns of
the data.frame (described in detail below). Adding one-to-many relationships within this
structure is trivial, as additional unviersalmotif data.frames can be nested
within eachother. These universalmotif data.frames form the basis for a majority
of `memes` data outputs. These structures are also valid input types to `memes`
functions, and when used as such, output data are appended as new columns to the
input data, ensuring data provenance. Finally, because the universalmotif
data.frame is just a data.frame, it can be manipulated using base R or
`tidyverse` functions, just like regular R data.frames, lowering the cognitive
overhead to interfacing with these structures.

@@

universalmotif data.frames are not limited to storing motifs from MEME Suite
data, and can additionally be used to examine or manipulate the properties of
any motif list. Using `universalmotif` functions to import motif data,
the `memes` function `as_universalmotif_dataframe()` converts any `universalmotif` object
into a `universalmotif` data.frame.  Manipulation of linked metadata columns
will result in updating the linked `universalmotif` object when calling
`update_motifs()` to retain the data.frame structure, or when calling
`as_universalmotif()` to convert the data.frame back into `universalmotif`
format. In this way, `memes` enables large-scale motif database manipulation &
analysis within the preexisting R framework.

The universalmotif data.frames generated by memes are not limited to storing
motifs from MEME Suite data. They can also be used to manipulate or store other
descriptive properties of motifs from multiple sources. `universalmotif`
provides several import functions which can convert motifs from external
databases queried from the `MotifDb` package, or from files in a variety of
non-meme file formats \cite{motifdb}. Once imported in `universalmotif` format
the `memes` function `as_universalmotif_dataframe()` converts any
`universalmotif` object into a universalmotif data.frame. Similar to MEME Suite
data stored in this format, if changes are made to the linked metadata columns,
the linked universalmotif object can be updated by calling `update_motifs()` to
retain the data.frame structure for further analysis, or by calling
`as_universalmotif()` to convert the data.frame back into universalmotif format,
discarding columns which cannot be stored within `universalmotif` format.
Therefore, this data structure empowers users to import and analyze motif lists
from a diverse set of sources for use with `memes` or simply for manipulating
motif metadata using common R tools. In this way, memes enables large-scale
motif database manipulation & analysis within the preexisting R framework.

@@

Motif analysis is often employed in ChIP-seq analysis, in which data is stored
as genomic coordinates rather than sequence. However, MEME Suite tools are
designed to work with sequences. While existing tools such as bedtools can
extract DNA sequence from gneomic coordinates, some MEME tools require fasta
headers to be specifically formatted. As a result, users must write custom code
to extract the DNA sequence for their genomic ranges of interest.

@@

### Data-aware motif analysis

Discriminative (or "differential") motif analysis, in which motifs are discovered in one set of
sequences relative to another set, can uncover biologically relevant motifs
associated with membership to distinct categories. For example, during analysis
of multiomic data, users can identify different functional categories of genomic
regions through integration with orthogonal datasets (e.g.  categorizing
transcription factor binding sites by the presence or absence of another
factor). In these instances, motif analysis acts as a powerful tool to examine whether
differences in motif content is associated with membership of distinct
categories (e.g. motifs enriched at sites where both factors are bound vs singly
bound). Although the MEME Suite allows differential enrichment testing, it does
not inherently provide a mechanism for analyzing groups of sequences in
parallel, or performing motif analysis with any understanding of data
categories. The `memes` framework enables "data-aware" motif analysis workflows
by allowing named lists of `Biostrings` objects as input to each function. If
using `GenomicRanges`, users can split peak data on a metadata column using the
base R `split` function, and then use this result as input to `get_sequence`, which
will produce a list of `BioStrings` objects where each entry is named after the
data categories from the split column. When a list is used as input to a `memes`
function, it runs the corresponding MEME Suite tool for each object in the list.
Users can also pass the name(s) of a category to the `control` argument to
enable differential analysis of the remaining list members against the
control category sequences. Therefore, `memes` enables complex data-aware
differential motif analysis workflows using simple syntax to extend the
capabilities of the MEME Suite.

@@

The TomTom tool allows users to compare unknown motifs to a set of known motifs
to identify the best match. Visual inspection of motif comparison data is a key
step in assessing the quality and accuracy of a match. The `view_tomtom_hits`
function allows users to compare query motifs with the list of potential matches
as assigned by TomTom, similar to the commandline behavior. The
`force_best_match` function allows users to reassign the TomTom best match
to a lower-ranked (but still statistically significant) match in order to
highlight motifs with greater biological relevance (e.g. to skip over a
transcription factor that is not expressed in the experimental sample, or when
two motifs are matched equally well).
